---
title: "Economy"
author: "@VhinsInc."
---

# Economy 
## GDP of Gabon

Gabon, located on the west coast of Central Africa, has a relatively small but resource-rich economy. The country's GDP (Gross Domestic Product) has been largely driven by its oil sector, which accounts for a significant portion of government revenue and export earnings. Additionally, Gabon is also known for its natural resources such as manganese, timber, and agricultural products.

### GDP Growth

Over the years, Gabon has experienced varying GDP growth, often linked to fluctuations in global oil prices. In periods of high oil prices, the economy has seen considerable growth, while downturns in the oil market have led to slower economic growth or contraction.

### Historical GDP Trends

From 1960 to 2023, Gabon's GDP has seen significant shifts. The country's economy grew rapidly in the 1970s and 1980s due to the discovery of oil, but growth has since slowed. While oil remains the dominant sector, diversification efforts have been made in recent years to expand into other sectors such as agriculture, mining, and services.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(WDI)
library(ggplot2)
library(data.table)
library(plotly)
## GDP
gabon_GDP <- WDI(country = "GA", indicator = "NY.GDP.MKTP.CD", start = 1960, end = 2023)


setDT(gabon_GDP)
setnames(gabon_GDP, "NY.GDP.MKTP.CD", "GDP")

setorder(gabon_GDP, year)

# gabon_GDP[, GDP := GDP]


gabon_GDP[, variation := round((GDP - shift(GDP, 1)) / shift(GDP, 1) * 100, 2)]


gabon_GDP[, second_variation := round((variation - shift(variation, 1)) / shift(variation, 1) * 100, 2)]


gabon_GDP[, sign_change := sign(second_variation) != shift(sign(second_variation), 1, type = "lag")]

change_years <- gabon_GDP[sign_change == TRUE, year]


gabon_GDP$second_derivative_color <- ifelse(gabon_GDP$second_variation < 0, "red", "green")



ggplot_obj <- ggplot(gabon_GDP, aes(x = year, y = GDP)) + 
  geom_line(color = "steelblue", size = 1) +  # Line plot
  geom_point(color = "red", size = 2) +        # Points for each year
  geom_vline(data = gabon_GDP[sign_change == TRUE], aes(xintercept = year), 
             color = "black", linetype = "dashed", size = 1) +  # Vertical lines where acceleration changes sign
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "green", se = FALSE) +  # Polynomial smoothing line
  scale_y_continuous(labels = scales::comma) + # Format Y-axis with commas
  labs(
    title = "Evolution of Gabon's GDP",
    x = "Year",
    y = "GDP [USD]",
    caption = "Source: World Bank"
  ) +
  theme_minimal() +  # Minimal theme for a clean look
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Centered bold title
    axis.title = element_text(size = 12),  # Axis titles size
    axis.text = element_text(size = 10)    # Axis text size
  )


plotly_obj <- ggplotly(ggplot_obj)

plotly_obj
```

### Economic Challenges

Despite its natural wealth, Gabon faces several economic challenges: - **Dependence on Oil**: The heavy reliance on oil exports makes the economy vulnerable to price fluctuations on the global market. - **Diversification**: There is an ongoing need for economic diversification to reduce reliance on oil and ensure sustainable growth. - **Income Inequality**: While the country has high GDP per capita compared to many other African nations, wealth distribution remains uneven.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot_obj1 <- ggplot(gabon_GDP, aes(x = year, y = variation)) + 
  geom_line(color = "steelblue", size = 1) +  # Line plot
  geom_point(color = "red", size = 2) +        # Points for each year
  geom_vline(data = gabon_GDP[sign_change == TRUE], aes(xintercept = year), 
             color = "black", linetype = "dashed", size = 1) +  # Vertical lines where acceleration changes sign
  # Shading the area based on the second derivative
  scale_y_continuous(labels = scales::comma) + # Format Y-axis with commas
  labs(
    title = "GDP growth in Gabon (2000-2023)",  # Updated title
    x = "Year", 
    y = "GDP growth (%)",  # Adjusted Y-axis label
    caption = "Source: World Bank"
  ) + 
  theme_minimal() +  # Minimal theme for a clean look
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Centered bold title
    axis.title = element_text(size = 12),  # Axis titles size
    axis.text = element_text(size = 10)    # Axis text size
  )


ggplot_obj1 <- ggplotly(ggplot_obj1)
ggplot_obj1

```

### Conclusion

Gabon's GDP provides a snapshot of a country with significant natural resources but also one that must navigate challenges related to diversification and global economic shifts. As the world moves toward renewable energy and more sustainable practices, Gabon's future economic trajectory may depend on how well it adapts to these changes.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

gdp_var <- na.omit(gabon_GDP$variation)


library(fitdistrplus)
library(MASS)



# Fit different distributions
fit_norm <- fitdistr(gdp_var, "normal")  # Normal distribution
fit_t <- fitdistr(gdp_var, "t")    



# Create a sequence of values covering the range of GDP variation
x_vals <- seq(min(gdp_var), max(gdp_var), length.out = 100)

# Compute fitted densities
dens_norm <- dnorm(x_vals, mean = fit_norm$estimate["mean"], sd = fit_norm$estimate["sd"])
dens_t <- dt((x_vals - fit_t$estimate["m"])/fit_t$estimate["s"], df = fit_t$estimate["df"]) / fit_t$estimate["s"]  # Adjusted Student’s t-density

# Store in a data.table
density_data <- data.table(
  x = rep(x_vals, 2),
  density = c(dens_norm, dens_t),
  distribution = rep(c("Normal", "Student's t"), each = length(x_vals))
)

# Plot empirical density with fitted distributions


ggplot_obj1 <- ggplot() +  
  # Empirical Density (Gray Fill)  
  
  # Fitted Distributions with Fill  
  geom_area(data = density_data, aes(x = x, y = density, fill = distribution), alpha = 0.3, position = "identity") +  
  geom_line(data = density_data, aes(x = x, y = density, color = distribution), size = 1) +   
  geom_density(data = gabon_GDP, aes(x = variation), fill = "gray", color = "black", alpha = 0.4) +   
  
  # Custom Colors  
  scale_fill_manual(values = c("Normal" = "blue", "Student's t" = "red")) +  
  scale_color_manual(values = c("Normal" = "blue", "Student's t" = "red")) +  
  
  # Labels & Theme  
  labs(
    title = "Comparison of Fitted Distributions to GDP Growth Variation in Gabon",
    x = "GDP Growth (%)",
    y = "Density",
    caption = "Source: World Bank"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_blank(),
    legend.position = "top"
  )

# Convert to interactive plot  
ggplot_obj1 <- ggplotly(ggplot_obj1)  
ggplot_obj1
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Generate 5 future points from the fitted normal distribution
set.seed(123)  # For reproducibility
years <- 14
future_variation <- rnorm(years, mean = fit_norm$estimate["mean"], sd = fit_norm$estimate["sd"])
future_variation <- rt(years, df = fit_t$estimate["df"]) * fit_t$estimate["s"] + fit_t$estimate["m"]


# Assign future years (2024–2028)
future_data <- data.table(
  year = max(gabon_GDP$year) + 1:years,
  variation = future_variation,
  simulated = TRUE  # Flag to differentiate real vs simulated data
)



# Compute Future GDP Sequentially
last_known_GDP <- tail(gabon_GDP$GDP, 1)  # Get last recorded GDP
future_data[, GDP := last_known_GDP * cumprod(1 + variation / 100)]  # Apply growth rates



# Add simulated data to the original dataset
# gabon_GDP_extended <- rbind(gabon_GDP[, .(year, variation, simulated = FALSE)], future_data)
gabon_GDP_extended <- rbind(
  gabon_GDP[, .(year, variation, GDP, simulated = FALSE)], 
  future_data
)



# Plot GDP variation sequence
ggplot_obj2 <- ggplot(gabon_GDP_extended, aes(x = year, y = variation)) +
  geom_line(color = "steelblue", size = 1) +  # Line connecting points
  geom_point(aes(color = simulated), size = 3) +  # Points (colored by real vs simulated)
  
  # Custom Colors
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "green"), 
                     labels = c("Historical Data", "Simulated Future Data")) +
  
  # Labels & Theme
  labs(
    title = "GDP Growth Variation in Gabon (Historical & Forecasted)",
    x = "Year",
    y = "GDP Growth (%)",
    caption = "Source: World Bank | Green: Simulated values (2024-2028)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_blank(),
    legend.position = "top"
  )

# Convert to interactive plot
ggplot_obj2 <- ggplotly(ggplot_obj2)
ggplot_obj2
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}


ggplot_obj3 <- ggplot(gabon_GDP_extended, aes(x = year, y = GDP)) +
  geom_line(color = "steelblue", size = 1) +  # Line connecting points
  geom_point(aes(color = simulated), size = 3) +  # Points (colored by real vs simulated)
  
  # Custom Colors
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "green"), 
                     labels = c("Historical GDP", "Projected GDP")) +
  
  # Labels & Theme
  labs(
    title = "GDP Evolution in Gabon (Historical & Forecasted)",
    x = "Year",
    y = "GDP (USD)",
    caption = "Source: World Bank | Green: Simulated GDP (2024-2028)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_blank(),
    legend.position = "top"
  )


ggplot_obj3 <- ggplotly(ggplot_obj3)
ggplot_obj3
```
