---
title: ""
---

## Electricity Infrastructure in Gabon

This map displays the **electricity infrastructure** in Gabon, including both the **electrical lines** and the **pylon locations**. The data is sourced from OpenStreetMap (OSM) and includes the following key elements:

### Key Features:

-   **Electricity Lines**: Represented as red lines on the map, these indicate the various electrical power transmission lines across the country. The lines are color-coded to represent their voltage level. Hovering over the lines will show the **voltage level** in volts (V).

-   **Pylons**: These blue circle markers show the locations of pylons. These pylons are essential for the transmission and distribution of electrical power. When you click on a marker, you will see the label "Crossing", which refers to the associated pylon crossing.

### Map Features:

-   **Interactive Layers**: The map allows users to interact with both the electricity lines and pylons, providing detailed information when clicked.

-   **Popup Information**: Each feature on the map (both the lines and pylons) is linked to a popup that displays relevant information about the element, such as the voltage of the electricity line or a simple label for pylons.

-   **Layer Control**: The map includes an interactive layers control option, allowing users to toggle between different features, such as electricity lines and pylons.

This map provides valuable insights into the **electricity infrastructure** of Gabon, making it easier to understand the layout and the critical locations of energy transmission across the country.

```{r, echo=FALSE, warning=FALSE}
library(leaflet)
library(sf)

path.osm <- "/Users/vhinymombo/Downloads/202303_OSM2IGEO_GABON_SHP_WGS84_4326/C_OSM_TRANSPORT_ENERGIE/"



# Charger les voies ferrées (lignes)
gabon_elec_sf <- st_read(path.osm, layer = "LIGNE_ELECTRIQUE", quiet = TRUE)

# Charger les passages à niveau (points)
gabon_pylone <- st_read(path.osm, layer = "PYLONE", quiet = TRUE)

voltage_pal <- colorNumeric(palette = c("red", "yellow", "green"), domain = gabon_elec_sf$VOLTAGE)

# Créer la carte avec les voies ferrées
masp <- leaflet() %>%
  addTiles() %>%
  addProviderTiles(providers$OpenStreetMap, group = "Open Street Map") %>% 
  addProviderTiles(providers$Esri.WorldImagery, group = "Esri World Imagery") %>% 
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Esri World Topography") %>% 
  
  
  addPolylines(
    data = gabon_elec_sf,
    color = ~voltage_pal(VOLTAGE),  # Apply the color palette based on VOLTAGE
    weight = 2,
    opacity = 0.8,
    popup = ~paste("VOLTAGE:", VOLTAGE, " V"),
    group = "Line"
  ) %>%
  addCircleMarkers(
    data = gabon_pylone,
    radius = 1,
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.8,
    popup = ~paste("Pylone"),
    group = "Pylone"
  ) %>%
  # Add layer control
  addLayersControl(
    options = layersControlOptions(collapsed = TRUE),
    overlayGroups = c("Pylone", "Line"),
    baseGroups = c("Esri World Imagery", "Open Street Map", "Esri World Topography"), # Tile providers
    
  ) %>%
    addLegend(
    position = "bottomright",  # Adjust position of the legend
    pal = voltage_pal,         # Use the defined color palette
    values = gabon_elec_sf$VOLTAGE,  # The values corresponding to the palette
    title = "Voltage (V)",     # Title of the legend
    opacity = 1                # Set opacity for the legend
  )



# Afficher la carte finale
masp

```

```{r}

```
